Perfomance Factors of Deep Learning and Shallow Neural Network Applications: A Beginners Guide

### 2018-01-12 paper.tex
again some changes

Machine Learning is probably the most present topic of computer science in general media. This popularity only increased over the recent years thorugh widely publicly known applications ranging from DeepBlue to IBM Watson and AlphaGo. Because of these successes Machine Learning is sometimes viewed as an omnipotent solution to every imaginable problem. This is most certainly not the case, based on our studies we can conclude that their capabilities depend on done design and engineering work. In this paper we want to give an short introduction into the most performance influencing factors to performance, during the most time intensive phase, the training.

\end{abstract}

@@

During our study we found no comprehensive study touching this thema not being based on the comparison of frameworks or on an experimental execution on different platforms or frameworks. This paper is based on these publications.

 At first we attend to the choice of neural network and the impact of this decision. Afterwards we have a look, on one hand, at small scale neural networks running on a single machine without large data sets and on the other hand, large scale neural networks operating with more layers and nodes, large data sets and a distributed system to handle the computational power required by the neural network training algorithms to be completed in reasonable time. We divide our survey into, the choice of neural network, the choice of processing units, including basic data distribution strategies for multi-processing, a discussion of the challenges in running small and large scales.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{neuralnetworks.png}
\caption{"A mostly complete chart of Neural Networks" by "The Asimov Institute" extract of only in this paper referenced networks\cite{zoo}}
\end{figure}

@@

The term "Deep Learning" itself is just a rebranding of the since the 80s existing concept of multilayered perceptrons. Its basic idea is to simply use more than 1 layer in a neural network to increase the accuracy of the learned model\cite{deeplearning101}. A good defintion of Deep Learning is delivered by Deng in his 2014 book "Deep learning: methods and applications"\cite{deng2014deep}: "A class of machine learning techniques that exploit many layers of non-linear information processing for supervised or unsupervised feature extraction and transformation, and for pattern analysis and classification". Additionally to the reuse of multilayer perceptrons the ideas of Deep Belief Networks and unsupervised pretraining have been introduced.

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{pop.png}
\caption{Popularity in \% of the term "deep learning" on Google in relation to maximum peak, plotted with the Google Trends tool provided by Google}
\end{figure}

Since then Deep Learning has experienced an almost exponential growth in popularity as seen in Figure 2, where the popularity steeply increases around 2014.
%%%%END DEEP LEARNING

@@

%simple
The most commonly used shallow Neural Networks are Feed Forward, Auto Encoders and Restricted Boltzmann Machines, which all have only a single layer of hidden nodes and therefore are quite restricted in what they can compute based on the input data.

Feed Forward Networks(FFN) simply connects input, hidden and output layers with an all to all connection between the layers with the number of outputs being equal or smaller to the number of hidden nodes.

Auto Encoders(AE) are of an similar build but the hidden layer compresses the input information into fewer nodes while these are later again decompressed to output nodes that match the previously given input nodes, therefore AE can be classified as a generative network.

@@

%deep
A more advanced architecture is to increase the number of hidden layers to enlarge the application possibility of these networks for more complex models these are then called Deep Networks. The more commonly used ones are Variational Auto Encoders , Deep Belief Networks, Generative Adversial Networks, Recurrent Neural Networks ,e.g. Long short-term memory networks, Deep Convolutional Neural Networks, Deconvolutional Networks, and Recursive Neural Networks.

@@

Deep Convolutional Networks(DCN) are based on Convolutional Nodes compressing the incoming information into the needed base model and then calculating the required output.\cite{PattersonGibson17} They are used to abstract input information into more general models. These networks especially are being used to process images.

@@

%%%%BEGIN TRAINING ALGORITHMS
%%THE POSITION OF THIS CAN BE CHANGED
The reason why the training of neural networks is so expensive can be traced back to the algorithm used to train all commonly used networks, which is backpropagation. Backpropagation is based on the idea that we first calculate the deviation of the output values from the actual result and then use this backwards thorugh the layers of the net to calculate a correction,based on the error of the node and the learning rate, for each connection weight until the input layer is reached\cite{riedmiller1993direct}. Some Networks like DBN use a slightly 
diverged "Gentle" Backpropagation, which relies on a learning rate after the initial values have been found from the training data with the help of the pretrained RBM \cite{PattersonGibson17}.
%%%%END TRAINING ALGORITHMS

@@

Further can the required training data exceed direct accessible memory from the processing unit, enhancing the problem of delivering data to all processing units.

@@

The next major performance influencing factor will be the choice of the actual processing unit. This decision is the most computation time influencing. This decision can have the biggest influence on computation time. We divide our processing unit into three groups starting with the CPU,which have a greater memory capability because they can be equipped with additional RAM sticks but possess a smaller parallelization limitation than GPUs because of their different architecture, then the GPU,with a much larger architectural focus on multithreading than CPUs, and following GPU Clusters, being simply a group of GPUs either in a single node or spanning multiple nodes, like they are used in supercomputers.

@@

%Transition
Next will follow the choice of neural network and its influence on performance.

@@

%%%%BEGIN ADVANTAGES DISADVANTAGES SIMPLE AND DEEP
\subsection{Choice of Neural Network}
\subsubsection{Shallow Networks}
Simple task can be done by a network which is not deep, like a FFN or a RBM. They are faster to train and even larger training sets can be completed in a reasonable time without multithreading or usage of GPUs. But their limited depth can decrease the accuracy of the model when the learned task is more complex than expected. In other words, these models have a risk to overfit the training data. Accordingly, to use them, careful validation is required.

\subsubsection{Deep Networks}
More complex tasks often require a deep neural network that can learn a more complex model than shallow networks and offer a greater accuracy while doing this. The major downside of these networks is that they require a larger amount of time to train and can be oversized for the actual learning task and might adding unnecessary computation time. Large scale applications mostly contain deep networks because of their larger training sets and more complex models underlying the actual task.

We can not offer a clear answer about when to use which kind of network, every problem must be individually analyzed and based on the information, and complexity, as well as the acceptable error rate of the final network the decision has to be made. For cases that are unclear a shallow network can first be constructed,because of their short training time, and if the result is unsatisfying a more complex network can be considered.  Also it is advisable to use a deep network if the learning task involves image or audio understanding, or the task needs to include recurring changes over time of an input signal\cite{PattersonGibson17}.
%%%%END ADVANTAGES DISADVANTAGES SIMPLE AND DEEP

@@

An also important choice is the setting of the initial values of connection weights, here it is advisable to start at low values like $0.1$ to prevent straying too far from the optimal solution, since algorithm like backpropagation mostly find a local optimum initial values too far of from the global optimum will deliver worse results.

@@

\subsubsection{Synchronization between GPUs}
For synchronizing Multi-GPU setups like in Figure 4 there are theoretically two choices. One being the synchronous method in which after a value is calculated it is immediately shared with every other GPU. This method is applicable to all Multi-GPU configurations and delivers an acceptable result\cite{wang2016deep}. The update of values for all GPUs occurs only after the updated values have been sent by each of them. Until a large amount of over 40 GPUs are used this mehtod can be used without limitation\cite{sastre2017scalability}. 
There is also the possibility to run the GPUs asynchronously, by letting them use different versions of the values and only updating them periodically by sending an update request. This can lead to a slight improvement of the training speed but it might also have a slight detrimental effect in the learning process\cite{wang2016deep}. This depends on the network model and training algorithm.

@@

\subsubsection{Multiple GPU-Clusters}
Taking a step further it is possible to connect multiple Clusters of GPUs to compute a single task. In doing this we again increase raw computation power, but also increase the total amount of communication required. Similar to multiple GPUs this leads at the beginning with 2 or more Clusters to a stronger improvement than later on with 8 to 16 Clusters as seen in (Figure 3)

>>>>

Machine Learning is probably the most present topic of computer science in general media. This popularity only increased over the recent years thorugh widely publicly known applications ranging from DeepBlue to IBM Watson and AlphaGo. Because of these successes Machine Learning, and the specialized sub-field of Deep Learning,  is sometimes viewed as an omnipotent solution to every imaginable problem. This is most certainly not the case. Based on our studies we can conclude that the capabilities of neural networks (i.e., the model used in deep learning) depend on the quality of their training. This process, in turn, can be time-consuming. Its performance depends on good design and engineering work. In this paper we want to give an short introduction into the most performance influencing factors when working with neural networks, with a focus on their training stage.
\end{abstract}

@@

During our research we found no comprehensive study touching this topic. The most closely related work are studies with a focus on comparing different platforms or frameworks. This paper is based on these publications.

 At first we attend to the choice of neural network and the impact of this decision. Afterwards we have a look, on one hand, at small scale neural networks running on a single machine without large data sets and on the other hand, large scale neural networks operating with more layers and nodes, large data sets and a distributed system to handle the computational power required by the neural network training algorithms to be completed in reasonable time. We divide our survey into
\begin{itemize}
\item the choice of neural network
\item the choice of processing units, including basic data distribution strategies for multi-processing
\item a discussion of the challenges in running small and large scales.
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{neuralnetworks.png}
\caption{"A mostly complete chart of Neural Networks" by "The Asimov Institute" extract of only in this paper referenced networks\cite{zoo}}
\end{figure}

@@

The term "Deep Learning" itself is just a rebranding of the since the 50s existing concept of multilayered perceptrons\cite{frank1957perceptron}. Its basic idea is to simply use more than 1 layer in a neural network to increase the accuracy of the learned model\cite{deeplearning101}. A good defintion of Deep Learning is delivered by Deng in his 2014 book "Deep learning: methods and applications"\cite{deng2014deep}: "A class of machine learning techniques that exploit many layers of non-linear information processing for supervised or unsupervised feature extraction and transformation, and for pattern analysis and classification". Additionally to the reuse of multilayer perceptrons the ideas of Deep Belief Networks and unsupervised pretraining were introduced in 2006\cite{hinton2006fast}, opening the door for continuous improvements in different aspects of neural networks. Among the relatively recent developments in neural networks has been the adoption of GPUs for processing, as exemplified in work winning the ImageNet challenge\cite{krizhevsky2012imagenet}.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{pop.png}
\caption{Popularity in \% of the term "deep learning" on Google in relation to maximum peak, plotted with the Google Trends tool provided by Google}
\end{figure}

Since then Deep Learning has experienced an almost exponential growth in popularity as seen in Figure 2, where the popularity steeply increased around 2014.
%%%%END DEEP LEARNING

@@

%simple
The most commonly used shallow Neural Networks are Feed Forward, Auto Encoders and Restricted Boltzmann Machines, which all have only a single layer of hidden nodes and therefore are quite restricted in what they can compute based on the input data. They are shown in Figure 1.

Feed Forward Networks(FFN) simply connect input, hidden and output layers with an all to all connection between the layers with the number of outputs being equal or smaller to the number of hidden nodes.

Auto Encoders(AE) are of a similar build but the hidden layer compresses the input information into fewer nodes while these are later again decompressed to output nodes that match the previously given input nodes, therefore AE can be classified as a generative network.

@@

%deep
A more advanced architecture is to increase the number of hidden layers to enlarge the application possibility of these networks for more complex models, these are then called Deep Networks. The more commonly used ones are Variational Auto Encoders , Deep Belief Networks, Generative Adversial Networks, Recurrent Neural Networks, e.g. Long short-term memory networks, Deep Convolutional Neural Networks, Deconvolutional Networks, and Recursive Neural Networks. They are also shown in Figure 1.

@@

Deep Convolutional Networks(DCN) are based on Convolutional Nodes compressing the incoming information into the needed base model and then calculating the required output\cite{PattersonGibson17}. They are used to abstract input information into more general models. These networks are specially being used to process images.

@@

%%%%BEGIN TRAINING ALGORITHMS
%%THE POSITION OF THIS CAN BE CHANGED
The reason why the training of neural networks is so expensive can be traced back to the algorithm used to train all commonly used networks, which is backpropagation. Backpropagation is based on the idea that we first calculate the deviation of the output values from the actual result and then use this backwards thorugh the layers of the net to calculate a correction,based on the error of the node and the learning rate, for each connection weight until the input layer is reached\cite{riedmiller1993direct}. Some Networks like DBN use a slightly 
different "Gentle" Backpropagation, which relies on a learning rate after the initial values have been found from the training data with the help of the pretrained RBM \cite{PattersonGibson17}.
%%%%END TRAINING ALGORITHMS

@@

@@

The next major performance influencing factor will be the choice of the actual processing unit. This decision is the most computation time influencing. This decision can have the biggest influence on computation time. We divide our processing unit into three groups starting with the CPUs, which have a greater memory capability because they can be equipped with additional RAM sticks but possess a smaller parallelization limitation than GPUs because of their different architecture, then the GPUs,with a much larger architectural focus on multithreading than CPUs, and following GPU-Clusters, being simply a group of GPUs either in a single node or spanning multiple nodes, like they are used in supercomputers.

@@

%Transition
Next will follow the choice of neural netwok as well as the setting of some hyper-parameters like the amount of nodes, followed by the choice of the processing unit

@@

%%%%BEGIN ADVANTAGES DISADVANTAGES SIMPLE AND DEEP
\subsection{Choice of Neural Network}
\subsubsection{Shallow Networks}
Simple learning task can be done by a network which is not deep, like a FFN or a RBM. They are faster to train and even larger training sets can be completed in a reasonable time without multithreading or usage of GPUs. But their limited depth can decrease the accuracy of the model when the learned task is more complex than expected. In other words, these models have a risk to overfit the training data. Accordingly, to use them, careful validation is required.

\subsubsection{Deep Networks}
More complex tasks often require a deep neural network that can learn a more complex model than shallow networks and offer a greater accuracy while doing this. The major downside of these networks is that they require a larger amount of time to train and can be oversized for the actual learning task, adding unnecessary computation time. Large scale applications mostly contain deep networks because of their larger training sets and more complex models underlying the actual task.

We can not offer a clear answer about when to use which kind of network. Every problem must be individually analyzed and based on the information, and complexity, as well as the acceptable error rate of the final network the decision has to be made. For cases that are unclear a shallow network can first be constructed,because of their short training time, and if the result is unsatisfying a more complex network can be considered.  Also it is advisable to use a deep network if the learning task involves image or audio understanding, or the task needs to include recurring changes over time of an input signal\cite{PattersonGibson17}.
%%%%END ADVANTAGES DISADVANTAGES SIMPLE AND DEEP

@@

Initial values, next to the number of nodes, the type of network, the activation functions, the activation thresholds, learning rate, the size of the batches, the training algorithm, the number of training epochs, and others, are called the hyper-parameters of the network. We can define these, informally, as the configurable factors that determine the training process in the network and the convergence of error rates. In our discussion, (given that our focus is on performance, with a data management perspective) we only consider the number of nodes. The precise effect of other hyper-parameters is, arguably less data management related, and thus, beyond the scope of our work.

@@

\subsubsection{Synchronization between GPUs}
For synchronizing Multi-GPU setups like in Figure 5 there are theoretically two choices. One being the synchronous method in which after a value is calculated it is immediately shared with every other GPU. This method is applicable to all Multi-GPU configurations and delivers an acceptable result\cite{wang2016deep}. The update of values for all GPUs occurs only after the updated values have been sent by each of them. Until a large amount of over 40 GPUs are used this mehtod can be used without limitation\cite{sastre2017scalability}. 
There is also the possibility to run the GPUs asynchronously, by letting them use different versions of the values and only updating them periodically by sending an update request. This can lead to a slight improvement of the training speed but it might also have a slight detrimental effect in the learning process\cite{wang2016deep}. This depends on the network model and training algorithm.

@@

\subsubsection{Multiple GPU-Clusters}
Taking a step further it is possible to connect multiple Clusters of GPUs to compute a single task. In doing this we again increase raw computation power, but also increase the total amount of communication required. Similar to multiple GPUs this leads at the beginning with 2 or more Clusters to a stronger improvement than later on with 8 to 16 Clusters as seen in (Figure 4)
